\Ace: An Actively-Typed Compilation Environment\\for High-Performance Computing

### 2012-10-14 ace/ace.tex
ace: Revised through section 2.G, part of 2.H

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\usepackage{listings}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape,
  basicstyle=\ttfamily\footnotesize,
  morekeywords={lambda, self, assert},
  numbers=left,
  numberstyle=\footnotesize\color{gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
}

@@

\subsection{Type Propagation and Higher-Order Functions}
The type assigned to the third argument, \verb|f|, on Lines 9 and 10, is given as \verb|add5.ace_type|. The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case. Generic functions are compiled automatically when they are called. That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type, and similarly with the \verb|int| type when compiling \verb|map_add5_int|. This mechanism is called {\em type propagation}. In other words, we did not need to run \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although doing so would also be valid). Only functions that are never called in the process of compiling other functions in a concrete module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map|.

We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not support function pointers.) This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as you would expect.) To put it in type theoretic terms, \verb|ace.GenericFnType|s are {\em uniquely inhabited} by a single generic function (and similarly for \verb|ace.ConcreteFnType|s), and thus it is impractical to use them as full first-class values (i.e. they cannot put into a run-time array.) In practice, this limitation is rarely a problem, particularly in parallel programming where such forms of specialization are already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.

We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or by using a jump table implementation, but we do not discuss this further due to lack of space. 
%%% TODO: Refer to line numbers in output

\subsection{Type Inference}
Within the \verb|map| function in Listing \ref{map}, on Line 5, the variable \verb|gid| is initialized with the result of calling the \verb|get_global_id| primitive (OpenCL supports multidimensional global ids, so the argument specifies which dimension to use.)  Note that a type for the \verb|gid| variable is not listed explicitly. This is because Ace supports a form of within-function {\em type inference}. In this case, \verb|gid| will have type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification, which the \verb|ace.OpenCL| module follows.) This can be seen on Lines 11 and 24 in Listing \ref{mapout}.

Inference is not restricted to single assignments, as in the \verb|map| example. Multiple assignments to the same variable with values of differing types and multiple return statements can be unified, such that the variable or return type is given a common supertype. For example, in Listing \ref{inference} the variable \verb|y| in the first branch of the conditional is assigned the \verb|int| literal \verb|0|. However, in the second branch of the loop, its type depends on the types of \verb|x| and \verb|scale|. We show two examples where this is the case on Lines 10 and 11. However, type inference correctly unifies these two types according to the C99 rules governing numeric types (which can be defined by the user, as we show in Section \ref{att}). This example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements were placed directly in the branches of the conditional.

@@

\subsection{Attribute and Extension Inference}
In addition to type annotations, OpenCL introduces additional syntactic burdens. Developers must annotate functions that meet the requirements to be called from a host with the \verb|__kernel| attribute. The \verb|Ace.OpenCL.OpenCL| backend is able to check these requirements and add this annotation automatically. Additionally, several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma|. The OpenCL backend can also  automatically infer many of these cases as well and add the appropriate \verb|#pragma| declaration, consistent with the principle of conciseness followed for the Ace design as a whole. An example of this can be seen in Listing \ref{mapout}.

\subsection{Metaprogramming in Ace}
\begin{codelisting}
\begin{lstlisting}[commentstyle=\color{mauve}]
from ace.OpenCL import OpenCL
import ast, ace.astx as astx

plus = OpenCL.fn.from_string("""
def plus(a, b):
    return a + b
""")

add5_ast = astx.specialize(plus.ast, b=5)
add5 = OpenCL.fn.from_ast(add5_ast)
\end{lstlisting}
\caption{Metaprogramming with Ace, showing how to construct generic functions from strings and abstract syntax trees, and how to manipulate syntax trees at compile time.}
\label{metaprogramming}
\end{codelisting}
Metaprogramming refers to the practice of writing programs that manipulate other programs. There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions. OpenCL itself in particular relies on code generation as a fundamental mechanism, partially justifying its lack of support for higher-order programming \cite{pyopencl}. Ace supports programmatic compilation and higher-order constructs, as described above, and language extension, which we describe below, so several use cases for this feature have been eliminated. However, we demonstrate cases where this form of metaprogramming is useful in the case study described in Section \ref{casestudy}.

An Ace function can be constructed from a string containing its source using the \verb|from_source| variant of \verb|fn|, as demonstrated in Listing \ref{metaprogramming} on Lines 4-7. It can also be constructed directly from a Python abstract syntax tree (AST), available via the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10. The AST here is generated programmatically by transforming the syntax tree of the \verb|plus| function such that the argument \verb|b| becomes a constant, \verb|5|. This transformation, called \verb|specialize|, as well as several others are distributed with Ace in the \verb|ace.astx| module for convenience.
\subsection{Direct Invocation from Python}\label{direct}

\begin{codelisting}
\begin{lstlisting}
import numpy as np
import ace.OpenCL.bindings as cl
from listing3 import map
from listing4 import add5 # or Listing 7

ctx = cl.Context.for_device(0, 0)
d_in = ctx.to_device(np.ones((1024,)))
d_out = ctx.alloc(like=d_in)
map(d_in, d_out, add5, 
    global_size=d_in.shape, local_size=(128,))
out = ctx.from_device(d_out)
\end{lstlisting}
\caption{A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer and  invocation of a generic function, \texttt{map}, as a kernel without compiling it first.}
\label{py}
\end{codelisting}

As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections. Python is already widely used by professional end-users as the high-level scripting language. It has support for executing code written in low-level languages in several ways. Developers can call into native libraries using its foreign function interface (FFI) or a library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL targeting nVidia GPUs \cite{pycuda}.

The OpenCL language is designed specifically for this workflow, in that it exposes the compiler directly an API, rather than as a separate executable. Developers generate OpenCL source code as strings, compile it programmatically, then execute it using the run-time APIs that OpenCL also provides. Python bindings to the OpenCL APIs have been developed in a module called \verb|pyopencl| \cite{pyopencl}.

Ace supports a similar workflow as an alternative to using \verb|acecc| as described above. Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions. An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 9-10. The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8. This wrapper associated type information with each buffer, and this information is used to implicitly compile \verb|map| as appropriate. That is, explicit calls to the \verb|compile| method, as we have been showing, are unnecessary when using this method of invocation. The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and the grouping of these threads (the {\em local size}).

By way of comparison, the same program written using the OpenCL C API directly is an orders of magnitude larger and correspondingly more complex. A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far. Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further -- the reader is referred to the language documentation for additional details.

>>>>

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\usepackage{listings}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape,
  basicstyle=\ttfamily\footnotesize,
  morekeywords={lambda, self, assert, as},
  numbers=left,
  numberstyle=\footnotesize\color{gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
}

@@

\subsection{Type Propagation and Higher-Order Functions}
The type assigned to the third argument, \verb|f|, on Lines 4.9 and 4.10, is given as \verb|add5.ace_type|. The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case. Generic functions are compiled automatically when they are called from another function. That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly with the \verb|int| type when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts). This mechanism is called {\em type propagation}. In other words, we did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although we could if we would like). Only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map| in Listing \ref{mapadd5dbl}.

We note that this scheme allows for a form of higher-order functional programming even when targeting languages, like OpenCL, that have no support for higher-order functions (OpenCL, unlike C99, does not even support function pointers). This works because the \verb|ace.GenericFnType| for one function, such as \verb|add5|, is not the same as the \verb|ace.GenericFnType| for a superficially similar function, such as \verb|add6| (defined as one would expect). To put it in type theoretic terms, \verb|ace.GenericFnType|s are singleton types, {\em uniquely inhabited} by a single generic function, and thus it is impractical to use them as full first-class values (i.e. they cannot put into a run-time array.) In practice, this is rarely a problem, particularly in parallel programming where explicit specialization is already common in order to avoid the potential performance penalty associated with a function pointer dereference per call.
We note that fully-featured first-class functions can be implemented when targeting a backend that supports them, such as C99, or simulated by using a jump table implementation in OpenCL, but we do not discuss this further due to lack of space.

\subsection{Type Inference}
On Line 5 in the generic \verb|map| function in Listing \ref{map}, the variable \verb|gid| is initialized with the result of calling the \verb|get_global_id| primitive (the argument, \verb|0|, is not important for our purposes.)  Note that the type for the \verb|gid| variable is never listed explicitly. This is because Ace supports a form of {\em whole-function type inference}. In this case, \verb|gid| will be assigned type \verb|size_t| because that is the return type of \verb|get_global_id| (as defined in the OpenCL specification \cite{opencl}, which the \verb|ace.OpenCL| module follows.) The result can be observed on Lines 11 and 24 in Listing \ref{mapout}.

Inference is not restricted within single assignments, as in the \verb|map| example. Multiple assignments to the same variable with values of differing types, or multiple return statements, can be unified such that the variable or return type is given a common supertype. For example, in the \verb|threshold_scale| function defined in Listing \ref{inference}, the variable \verb|y| in the first branch of the conditional is assigned the \verb|int| literal \verb|0|. However, in the second branch of the loop, its type depends on the types of \verb|x| and \verb|scale|. We show two choices for these types on Lines 10 and 11. However, type inference correctly unifies these two types according to OpenCL's C99-derived rules governing numeric types (which are defined by the user in the \verb|OpenCL| module, as we will describe in Section \ref{att}). We test this programmatically on Lines 12 and 13. Note that this example would also work correctly if the assignments to \verb|y| were replaced with \verb|return| statements (in other words, the return value of a function is treated as an assignable for the purpose of type inference).

@@

\subsection{Annotation and Extension Inference}
In addition to type annotations, OpenCL normally asks for additional annotations in a number of other situations.  Users can annotate functions that meet certain requirements to be callable from the host with the \verb|__kernel| attribute. The \verb|Ace.OpenCL.OpenCL| backend is able to check these requirements and add this annotation automatically. Several types (notably, \verb|double|) and specialized functions require that an OpenCL extension be enabled with a \verb|#pragma| when used. The OpenCL backend automatically detects many of these cases as well and adds the appropriate \verb|#pragma| declaration. An example of this feature can be seen on Line 5.1, where the use of the \verb|double| type triggers the insertion of an appropriate \verb|#pragma| automatically. Ace is designed to allow backends to observe the results of the type checking process to support this form of inference.

\subsection{Metaprogramming in Ace}
\begin{codelisting}
\begin{lstlisting}[commentstyle=\color{mauve}]
from ace.OpenCL import OpenCL
import ace.astx as astx

plus = OpenCL.fn.from_str("""
def plus(a, b):
    return a + b
""")

add5_ast = astx.specialize(plus.ast, b=5)
add5 = OpenCL.fn.from_ast(add5_ast)
\end{lstlisting}
\caption{[\texttt{listing7.py}] Metaprogramming with Ace, showing how to construct generic functions from both strings and abstract syntax trees, and how to manipulate syntax trees at compile-time.}
\label{metaprogramming}
\end{codelisting}
Metaprogramming refers to the practice of writing programs that manipulate other programs. There are a number of use cases for this technique, including domain-specific optimizations and code generation for programs with a repetitive structure that cannot easily be captured using available abstractions \cite{pyopencl}. OpenCL itself in particular relies on code generation as a fundamental mechanism, partially justifying its lack of support for higher-order programming. Ace supports programmatic compilation and higher-order constructs, as described above, and a flexible language extension mechanism, which we describe below, so several use cases for metaprogramming have been eliminated. However, cases where this form of metaprogramming include programmatic function specialization (Listing \ref{metaprogramming}) and modular simulation orchestration, described in Section \ref{clegans}.

On Lines 4-7 of Listing \ref{metaprogramming}, an Ace function is constructed from a string containing its source using the \verb|from_source| variant of the \verb|OpenCL.fn| method. It can also be constructed directly from an abstract syntax tree (AST), as implemented by the Python standard \verb|ast| package, using the \verb|from_ast| variant of \verb|fn|, demonstrated on Line 10. The AST here is generated programmatically by copying the syntax tree of the \verb|plus| function with the argument, \verb|b|, eliminated and its uses replaced with a constant, \verb|5|. This transformation, called \verb|specialize|, as well as several others are distributed in the \verb|ace.astx| module for convenience.
\subsection{Direct Invocation from Python}\label{direct}

\begin{codelisting}
\begin{lstlisting}
import numpy as np
import ace.OpenCL.bindings as cl
from listing3 import map
from listing4 import add5 # or from listing7

ctx = cl.Context.for_device(0, 0)
d_in = ctx.to_device(np.ones((1024,)))
d_out = ctx.alloc(like=d_in)
map(d_in, d_out, add5, 
    global_size=d_in.shape, local_size=(128,))
out = ctx.from_device(d_out)
\end{lstlisting}
\caption{[\texttt{listing8.py}] A full OpenCL program using the \texttt{Ace.OpenCL} Python bindings, including data transfer to and from a device and direct invocation of a generic function, \texttt{map}, as a kernel without explicit compilation.}
\label{py}
\end{codelisting}

As discussed in the Introduction, a common workflow for professional end-users involves the use of a high-level scripting language for overall workflow orchestration and small-scale data analysis and visualization, paired with a low-level language for performance-critical sections. Python is already widely used by professional end-users in the role of the high-level scripting language. It also has mature support for executing code written in low-level languages, in several ways. Developers can call into native libraries using its foreign function interface (FFI),  or by using a wrapper library like \verb|pycuda| for code compiled with CUDA, a proprietary language similar to OpenCL specifically targeting nVidia GPU hardware. 
Although CUDA's compilers are separate executables, the OpenCL language is tailored for this workflow, in that it exposes the compiler directly an API. The \verb|pyopencl| module exposes this API as well as the OpenCL memory management API to Python. With both \verb|pyopencl| and \verb|pycuda|, developers generate source code as strings, compile it programmatically, then execute it using the run-time APIs that each library provides \cite{pyopencl}.

Ace supports this workflow as an alternative to using the \verb|acecc| compiler as described above to generate  source code directly. For the OpenCL backend, these bindings are exposed as a wrapper on top of \verb|pyopencl| called \verb|Ace.OpenCL.bindings|. Both generic functions and concrete functions written for a backend that supports the direct execution interface (thus far, \verb|OpenCL| and \verb|C99|) can be called like regular Python functions. An example of this for the generic \verb|map| function defined in Listing \ref{map} is shown in Listing \ref{py}, with the call itself on Lines 9-10. The first two arguments to \verb|map| are OpenCL buffers, generated using a simplified wrapper to the \verb|pyopencl| APIs on Lines 7-8. This wrapper associated type information with each buffer, and this information is used to implicitly compile \verb|map| as appropriate. That is, explicit calls to the \verb|compile| method, as we have been showing, are unnecessary when using this method of invocation. The final two keyword arguments on Line 10 are parameters that OpenCL requires for execution that determine the number of threads (called the {\em global size}) and the grouping of these threads (the {\em local size}).

By way of comparison, the same program written using the OpenCL C API directly is an order of magnitude larger and correspondingly more complex. A full implementation of the logic of \verb|map| written using the \verb|pyopencl| bindings and metaprogramming techniques as described in \cite{pyopencl} is twice as large and significantly more difficult to comprehend than the code we have shown thus far. Not shown are several additional conveniences, such as delegated kernel sizing and \verb|In| and \verb|Out| constructs that can reduce the size and improve the clarity of this code further -- due to a lack of space, the reader is referred to the language documentation for additional details on these features.
